Exploring location codes to work out what the Other location for deaths in NHS Borders represent

### 2021-11-08 location_code_exploration.Rmd
expanded location exploration

```{r}
library(tidyverse)
library(phsmethods)
```

@@

As will become clear, there are lots of codes in the `Other` category; from manual review they mainly comprise health centres, clinics, GP surgeries (A,B,C), residential living institutions,  primary schools (L), nurseries &  colleges (W), high schools (Q).

@@

The list of location codes can be found on the [National Reference Files website](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/), with the relevant data dictionary at the [SMR file reference](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/SMR-reference-file-layouts.asp#Location%20Codes)

@@

```{r}
## download location codes
if (!file.exists("./downloaded_data/location_codes")) download.file(url = "http://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/Nat-Ref-Files/location?52573360", destfile = "./downloaded_data/location_codes", method = "curl")
```

@@

@@

```{r}
codes_hospital <- c("H")
codes_home <- c("N")
codes_care_home <- c("J", "K", "R", "S", "T", "U","V")
```

@@

## What does the first letter in the `location code` mean?

I suspect `B` is for Borders HB...

```{r}
locations_hb_borders_other %>% count(str_sub(location_code,1,1))

locations %>% anti_join(locations_hb_borders_other) %>% count(str_sub(location_code, 1, 1))

locations %>% anti_join(locations_hb_borders_other) %>% filter(str_sub(location_code, 1, 1) == "B") %>% count(is.na(postcode))

locations %>% anti_join(locations_hb_borders_other) %>% filter(str_sub(location_code, 1, 1) == "B")
```

>>>>

```{r}
library(tidyverse)
library(phsmethods)
library(readxl)  # for reading excel files
library(lubridate)  # for working with dates & times
library(janitor)
```

@@

As will become clear, there are lots of codes in the `Other` category.

A list of all locations and their associated codes are available on the [NRS website](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-background-information/code-lists-used-in-vital-events-statistics), specifically at [the following link: Institutions](https://www.nrscotland.gov.uk/files//statistics/vital-events/institution-codes-october-2021.xlsx).

@@

~~The list of location codes can be found on the [National Reference Files website](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/), with the relevant data dictionary at the [SMR file reference](https://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/SMR-reference-file-layouts.asp#Location%20Codes)~~
Note: see above for the NRS-provided list of institutions which also explains the type of institution and health board coding.

@@

```{r}
## download location codes
## this is for the National Reference File, but it's better to use the NRS-provided file instead
# if (!file.exists("./downloaded_data/location_codes")) download.file(url = "http://www.isdscotland.org/Products-and-Services/Data-Definitions-and-References/National-Reference-Files/Nat-Ref-Files/location?52573360", destfile = "./downloaded_data/location_codes", method = "curl")

institutions_file_url <- "https://www.nrscotland.gov.uk/files//statistics/vital-events/institution-codes-october-2021.xlsx"
institutions_file_path <- "./downloaded_data/institution-codes-october-2021.xlsx"

if (!file.exists(institutions_file_path)) download.file(url = institutions_file_url, destfile = institutions_file_path, method = "curl")
```

@@

institution_codes <-
  c("A - Admin Office",
     "B - Health Centre, GP Surgery",
     "C - Clinic Premises",
     "E - Extra-mural clinic",
     "H - NHS Hospital",
     "J - Joint User Hospital",
     "K - Contractual Hospital",
     "L - School",
     "M - Non NHS Maternity",
     "N - Non-Institutional, e.g. Domiciliary",
     "P - Prison",
     "Q - School",
     "R - Home for the Elderly",
     "S - Other Home",
     "T - Miscellaneous Premises",
     "V - Private Nursing Home, Private Hospital etc",
     "W - School") %>%
  tibble(institution_code_nrs = .) %>%
  separate(col = institution_code_nrs, into = c("institution_code_nrs","institution_type"), sep = " - ", remove = TRUE)

institution_hb_codes <- read_excel(path = institutions_file_path, sheet = 1, range = "A7:B21", col_names = c("hb_code","hb"))
institution_types <- read_excel(path = institutions_file_path, sheet = 1, range = "E7:G21", col_names = c("institution_type_code","institution_type","institution_type_detail"))

institution_types <-
  bind_rows(
    institution_types,
    institution_codes %>% rename(institution_type_code = institution_code_nrs) %>% anti_join(institution_types, by = "institution_type_code")
  )

institutions <- 
  read_excel(path = institutions_file_path, sheet = 1, skip = 28, col_types	= rep(c("text","date"), each = 3)) %>%
  clean_names %>%
  filter(nchar(location_code)==5) %>%  # extract only actual entries, ignoring all the headers & formatting rows
  mutate(
    institution_type_code = str_sub(location_code, 5, 5),
    hb_code = str_sub(location_code, 1, 1)
  ) %>% 
  left_join(institution_hb_codes, by = "hb_code") %>%
  left_join(institution_types %>% select(institution_type, institution_type_code), by = "institution_type_code")

@@

```{r}
codes_hospital <- c("H")
codes_home <- c("N")
codes_care_home <- c("J", "K", "R", "S", "T", "U","V")

institutions <-
  institutions %>%
  mutate(place_of_death = case_when(
    institution_type_code %in% codes_hospital ~ "Hospital",
    institution_type_code %in% codes_home ~ "Home & other non-institutional setting",
    institution_type_code %in% codes_care_home ~ "Care Home",
    TRUE ~ "Other"
  ))
```

@@

```{r}
## relatively new (2016 or later institutions)
institutions %>%
  filter(hb=="Borders" & place_of_death == "Other") %>%
  filter(year(location_closure_date) > 2016 | is.na(location_closure_date)) %>%
  filter(year(location_start_date) >= 2016) %>%
  View()

institutions %>%
  filter(hb=="Borders" & place_of_death == "Other") %>%
  filter(str_detect(location_name, pattern = regex("base|royal|raf|army|milit", ignore_case = TRUE))) %>%
  .$location_name

institutions %>%
  filter(hb=="Borders" & place_of_death == "Other") %>%
  filter(year(location_closure_date) > 2016 | is.na(location_closure_date)) %>%
  View()
```
